WASP-4\lowercase{b} Arrived Early for the TESS Mission

### 2018-12-18 paper/ms.tex
update ms.tex w/ latest numbers

\begin{abstract}
    The Transiting Exoplanet Survey Satellite (TESS) recently observed
    18 transits of the hot Jupiter WASP-4b.  The sequence of transits
    occurred $81.5 \pm 11.7$~seconds earlier than had been predicted,
    based on data stretching back to 2007.  This is unlikely
    to be the result of a clock error, because TESS observations of
    other hot Jupiters (WASP-6b, 18b, and 46b) are compatible with a
    constant period, ruling out an 81.5-second offset at the
    6.4$\sigma$ level.  The 1.3-day orbital period of WASP-4b appears
    to be decreasing at a rate of $\dot{P} = -12.6 \pm 1.2$
    milliseconds per year.  This might be caused by tidal orbital
    decay or apsidal precession, although both interpretations have
    shortcomings.  The gravitational influence of a third body is another
    possibility, although at present there is no evidence for such a
    body. Further observations are needed to confirm and
    understand the timing variation.
\end{abstract}

@@

\begin{figure}[t]
    \begin{center}
        \leavevmode
        \includegraphics[width=0.49\textwidth]{f3.pdf}
    \end{center}
    \vspace{-0.6cm}
    \caption{ {\bf TESS saw WASP-4b transit earlier than expected.}
      Both plots show the deviations between the observed and
      calculated transit times, where the calculation is based only on
      the pre-TESS data and assumes a constant period.  The blue bands
      depict the $\pm$$1\sigma$ credible interval of the predicted
      times.  {\it Top:} The full timing dataset spans 11 years. The
      darkest points correspond to the most precise data.  {\it
        Bottom:} Close-up of the TESS observations. The red band shows
      the average deviation of the TESS transits, which arrived $81.5 \pm
      11.7\ {\rm seconds}$ earlier than predicted.
        \label{fig:arrived_early}
    }
\end{figure}

@@

Figure~\ref{fig:times} shows the residuals with respect to the
constant-period model.  The best-fitting constant-period model has
$\chi^2 = 174$ and 61 degrees of freedom.  The best-fitting
decreasing-period model has $\chi^2 = 62.6$ and 60 degrees of freedom.
The best-fitting precession model has $\chi^2 = 64.4$ and 58 degrees
of freedom.  Either of the latter two models provides a much better
fit than the constant-period model.  The difference in $\chi^2$
between the first and second models corresponds to $p \approx
10^{-26}$.

The decreasing-period model provides a slightly better fit to the data
than the precession model.  It is favored by $\Delta \chi^2 = 1.8$,
and has two fewer free parameters.  A useful heuristic for model
comparison is the Bayesian Information Criterion (BIC),
\begin{equation}
  {\rm BIC} = \chi^2 + k\log n,
\end{equation}
where $k$ is the number of free parameters, and $n$ is the number of
data points. In this case, $n=62$.  The difference in the BIC between
the precession and decay models is $\Delta {\rm BIC} = {\rm BIC}_{\rm
  prec} - {\rm BIC}_{\rm quad} = 10$, corresponding to a Bayes factor
of $1.2\times 10^{4}$.  Likewise, the Akaike Information Criterion
favors the constant-period-derivative model by $\Delta {\rm AIC} =
5.8$.  Differences of this magnitude are traditionally deemed ``strong
evidence'' that one model is a better description of the data than the
other \citep{kass_bayes_1995}, although we prefer to reserve judgment
until more data can be obtained.

In the decreasing-period model, the period derivative is
\begin{equation}
\dot{P}
  = - (3.98 \pm 0.38)\times 10^{-10}
  = - 12.6 \pm 1.2~{\rm ms}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
For comparison, the best-fitting period derivative of WASP-12b is
$\dot{P} = -29 \pm 3\,{\rm ms}\,{\rm yr}^{-1}$
\citep{maciejewski_departure_2016,patra_2017}.  If both planets are
truly falling onto their stars, then WASP-4b is falling at about half
the rate of WASP-12b.  In the precession model, the best-fit
eccentricity is
\begin{equation}
  e = (1.83^{+ 1.88}_{- 0.75})\times10^{-3}
\end{equation}
The longitude of periastron advances by $\dot{\omega} =
14.0^{+4.9}_{-3.5}\ {\rm degrees}\,{\rm yr}^{-1}$, and the precession
period is $26^{+9}_{-7}$ years.  All of the best-fitting parameters
(the medians of the posterior distributions) and the 68\% credible
intervals are reported in Table~4.

@@

If instead the observed timing variation is just a small portion of an
apsidal precession cycle, then the orbital eccentricity is a few times
$10^{-3}$, and the full precession period is about 24 years.
\citet{ragozzine_probing_2009} calculated apsidal precession periods
for hot Jupiters, finding them to range between about 10 and 100
years. They highlighted that for many hot Jupiters, including WASP-4b,
the theoretical precession rate is dominated by the non-Keplerian
force due to the planet's tidal bulge.  Precession from general
relativity, the planet's rotational bulge, and the star's rotational
and tidal bulges contribute at the 10\% level at most.  Thus, a
measurement of the precession rate can be used to determine the
planet's Love number.  From their Equation 14, the implied Love number
for WASP-4b is
\begin{equation}
k_{2,{\rm p}} = 1.5^{+2.1}_{-1.1}.
\end{equation}
For comparison, the Love number of Jupiter is about 0.55
\citep{wahl_tidal_2016,ni_empirical_2018}, and a uniform density
sphere has $k_2 = 1.5$. The uncertainty in $k_{2,{\rm p}}$ for WASP-4b
is large because the eccentricity, reference time, and ${\rm
d}\omega/{\rm d}E$ have strongly correlated errors, and the measured
occultation times only barely constrain these parameters
(Figure~\ref{fig:future}).

The main problem with the apsidal precession hypothesis is to explain
why the eccentricity would be as large as $\sim$$10^{-3}$ despite
rapid tidal circularization.  For WASP-4, Equation~\ref{eq:de_dt}
gives $\tau_e = 0.30 (Q_{\rm p}'/10^5) \, {\rm Myr}$.  The star is
several billion years old, so unless the planet arrived very recently,
any initial eccentricity should have been lowered well below
$\sim$$10^{-3}$.  The top panel of Figure~\ref{fig:context} compares
the expected eccentricity damping time of WASP-4b with that of other
transiting giant planets.  WASP-4b has one of the shortest known
eccentricity damping times.

@@

A different way to excite the hot Jupiter's eccentricity is the
Kozai-Lidov mechanism \citep{lidov_evolution_1962,kozai_secular_1962}.
In this case, the orbital plane of the outer companion, ``c'', would
need to be inclined relative to that of the hot Jupiter, ``b'', by at
least $\sin^{-1} \sqrt{2/5} \approx 39^\circ$.
For the Kozai-Lidov mechanism to operate at maximum efficiency, we
need \citep[][Equation 20]{bailey_understanding_2019}
\begin{equation}
  M_{\rm c} > 7.1\,M_\oplus
  \times \left( \frac{a_{\rm c}}{a_{\rm b}} \right)^{3/2},
  \label{eq:kozai_bound}
\end{equation}
where $M_{\rm b,c}$ and $a_{\rm b,c}$ are the mass and semi-major axis
of WASP-4b and the hypothetical WASP-4c, and we have assumed WASP-4b's
Love number is $k_{2,{\rm b}}\approx 0.6$.  For the RV signal of the
companion to remain undetected, it would need to be in the residual
$({\rm O-C})_{\rm RV} = 15.2\,{\rm m s}^{-1}$ reported by
\citet{triaud_spin-orbit_2010}.  Again following
\citet{bailey_understanding_2019}, this implies
\begin{align}
  M_{\rm c} &<
  ({\rm O-C})_{\rm RV}
  \left( \frac{M_\star a_b}{G} \right)^{1/2}
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2}
  \nonumber
  \\
  M_{\rm c} &< 
  23.8\,M_\oplus
  \times 
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2},
  \label{eq:rv_bound}
\end{align}
for $f(e_{\rm c},\omega_{\rm c}, i_{\rm c}) \propto \sin^2 i_{\rm c}$
a geometric prefactor that depends on the argument of periastron
$\omega_{\rm c}$ and inclination $i_{\rm c}$ of the exterior companion
(\citealt{bailey_understanding_2019} Equation 23).  Since WASP-4b is
transiting, $f$ can be arbitrarily small, and both of the preceding
limits can be satisfied.

@@

For each planet, we asked: what fraction of the posterior for $t_{\rm
offset}$ is consistent with an offset worse than $81.5$ seconds?  For
WASP-4b, the answer is unsurprisingly 50\%.  For WASP-6b, the most
constraining object, about 1 sample in 2 million is consistent with
such a timing offset ($4.9\sigma$).  For WASP-18b, 1 in 103 samples
would be consistent with this timing offset ($2.3\sigma$), and in
WASP-46b, the limit is 1 in 49 samples ($2.0\sigma$).  For WASP-5b,
the predicted time is too imprecise to rule out timing offsets at the
necessary amplitude.  Multiplying the three independent probabilities
for WASP-5b, 6b, and 18b, we can rule out $t_{\rm offset} < -81.5\
{\rm seconds}$ at $6.4\sigma$, or about about 1 part in 11 billion.

>>>>

\begin{abstract}
    The Transiting Exoplanet Survey Satellite (TESS) recently observed
    18 transits of the hot Jupiter WASP-4b.  The sequence of transits
    occurred $81.6 \pm 11.7$~seconds earlier than had been predicted,
    based on data stretching back to 2007.  This is unlikely
    to be the result of a clock error, because TESS observations of
    other hot Jupiters (WASP-6b, 18b, and 46b) are compatible with a
    constant period, ruling out an 81.6-second offset at the
    6.4$\sigma$ level.  The 1.3-day orbital period of WASP-4b appears
    to be decreasing at a rate of $\dot{P} = -12.6 \pm 1.2$
    milliseconds per year.  This might be caused by tidal orbital
    decay or apsidal precession, although both interpretations have
    shortcomings.  The gravitational influence of a third body is another
    possibility, although at present there is no evidence for such a
    body. Further observations are needed to confirm and
    understand the timing variation.
\end{abstract}

@@

\begin{figure}[t]
    \begin{center}
        \leavevmode
        \includegraphics[width=0.49\textwidth]{f3.pdf}
    \end{center}
    \vspace{-0.6cm}
    \caption{ {\bf TESS saw WASP-4b transit earlier than expected.}
      Both plots show the deviations between the observed and
      calculated transit times, where the calculation is based only on
      the pre-TESS data and assumes a constant period.  The blue bands
      depict the $\pm$$1\sigma$ credible interval of the predicted
      times.  {\it Top:} The full timing dataset spans 11 years. The
      darkest points correspond to the most precise data.  {\it
        Bottom:} Close-up of the TESS observations. The red band shows
      the average deviation of the TESS transits, which arrived $81.6 \pm
      11.7\ {\rm seconds}$ earlier than predicted.
        \label{fig:arrived_early}
    }
\end{figure}

@@

Figure~\ref{fig:times} shows the residuals with respect to the
constant-period model.  The best-fitting constant-period model has
$\chi^2 = 174$ and 61 degrees of freedom.  The best-fitting
decreasing-period model has $\chi^2 = 62.6$ and 60 degrees of freedom.
The best-fitting precession model has $\chi^2 = 64.3$ and 58 degrees
of freedom.  Either of the latter two models provides a much better
fit than the constant-period model.  The difference in $\chi^2$
between the first and second models corresponds to $p \approx
10^{-26}$.

The decreasing-period model provides a slightly better fit to the data
than the precession model.  It is favored by $\Delta \chi^2 = 1.7$,
and has two fewer free parameters.  A useful heuristic for model
comparison is the Bayesian Information Criterion (BIC),
\begin{equation}
  {\rm BIC} = \chi^2 + k\log n,
\end{equation}
where $k$ is the number of free parameters, and $n$ is the number of
data points. In this case, $n=62$.  The difference in the BIC between
the precession and decay models is $\Delta {\rm BIC} = {\rm BIC}_{\rm
  prec} - {\rm BIC}_{\rm quad} = 10$, corresponding to a Bayes factor
of $1.1\times 10^{4}$.  Likewise, the Akaike Information Criterion
favors the constant-period-derivative model by $\Delta {\rm AIC} =
5.8$.  Differences of this magnitude are traditionally deemed ``strong
evidence'' that one model is a better description of the data than the
other \citep{kass_bayes_1995}, although we prefer to reserve judgment
until more data can be obtained.

In the decreasing-period model, the period derivative is
\begin{equation}
\dot{P}
  = - (4.00 \pm 0.38)\times 10^{-10}
  = - 12.6 \pm 1.2~{\rm ms}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
For comparison, the best-fitting period derivative of WASP-12b is
$\dot{P} = -29 \pm 3\,{\rm ms}\,{\rm yr}^{-1}$
\citep{maciejewski_departure_2016,patra_2017}.  If both planets are
truly falling onto their stars, then WASP-4b is falling at about half
the rate of WASP-12b.  In the precession model, the best-fit
eccentricity is
\begin{equation}
  e = (1.92^{+ 1.93}_{- 0.76})\times10^{-3}
\end{equation}
The longitude of periastron advances by $\dot{\omega} =
13.6^{+4.7}_{-3.6}\ {\rm degrees}\,{\rm yr}^{-1}$, and the precession
period is $27^{+10}_{-7}$ years.  All of the best-fitting parameters
(the medians of the posterior distributions) and the 68\% credible
intervals are reported in Table~4.

@@

If instead the observed timing variation is just a small portion of an
apsidal precession cycle, then the orbital eccentricity is a few times
$10^{-3}$, and the full precession period is about 27 years.
\citet{ragozzine_probing_2009} calculated apsidal precession periods
for hot Jupiters, finding them to range between about 10 and 100
years. They highlighted that for many hot Jupiters, including WASP-4b,
the theoretical precession rate is dominated by the non-Keplerian
force due to the planet's tidal bulge.  Precession from general
relativity, the planet's rotational bulge, and the star's rotational
and tidal bulges contribute at the 10\% level at most.  Thus, a
measurement of the precession rate can be used to determine the
planet's Love number.  From their Equation 14, the implied Love number
for WASP-4b is
\begin{equation}
k_{2,{\rm p}} = 1.5^{+2.1}_{-1.1}.
\end{equation}
For comparison, the Love number of Jupiter is about 0.55
\citep{wahl_tidal_2016,ni_empirical_2018}, and a uniform density
sphere has $k_2 = 1.5$. The uncertainty in $k_{2,{\rm p}}$ for WASP-4b
is large because the eccentricity, reference time, and ${\rm
d}\omega/{\rm d}E$ have strongly correlated errors, and the measured
occultation times only barely constrain these parameters
(Figure~\ref{fig:future}).

The main problem with the apsidal precession hypothesis is to explain
why the eccentricity would be as large as $\sim$$10^{-3}$ despite
rapid tidal circularization.  For WASP-4, Equation~\ref{eq:de_dt}
gives $\tau_e = 0.29 (Q_{\rm p}'/10^5) \, {\rm Myr}$.  The star is
several billion years old, so unless the planet arrived very recently,
any initial eccentricity should have been lowered well below
$\sim$$10^{-3}$.  The top panel of Figure~\ref{fig:context} compares
the expected eccentricity damping time of WASP-4b with that of other
transiting giant planets.  WASP-4b has one of the shortest known
eccentricity damping times.

@@

A different way to excite the hot Jupiter's eccentricity is the
Kozai-Lidov mechanism \citep{lidov_evolution_1962,kozai_secular_1962}.
In this case, the orbital plane of the outer companion, ``c'', would
need to be inclined relative to that of the hot Jupiter, ``b'', by at
least $\sin^{-1} \sqrt{2/5} \approx 39^\circ$.
For the Kozai-Lidov mechanism to operate at maximum efficiency, we
need \citep[][Equation 20]{bailey_understanding_2019}
\begin{equation}
  M_{\rm c} > 7.5\,M_\oplus
  \times \left( \frac{a_{\rm c}}{a_{\rm b}} \right)^{3/2},
  \label{eq:kozai_bound}
\end{equation}
where $M_{\rm b,c}$ and $a_{\rm b,c}$ are the mass and semi-major axis
of WASP-4b and the hypothetical WASP-4c, and we have assumed WASP-4b's
Love number is $k_{2,{\rm b}}\approx 0.6$.  For the RV signal of the
companion to remain undetected, it would need to be in the residual
$({\rm O-C})_{\rm RV} = 15.2\,{\rm m s}^{-1}$ reported by
\citet{triaud_spin-orbit_2010}.  Again following
\citet{bailey_understanding_2019}, this implies
\begin{align}
  M_{\rm c} &<
  ({\rm O-C})_{\rm RV}
  \left( \frac{M_\star a_b}{G} \right)^{1/2}
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2}
  \nonumber
  \\
  M_{\rm c} &< 
  23.7\,M_\oplus
  \times 
  \left( \frac{a_c}{a_b} \right)^{1/2}
  f^{-1/2},
  \label{eq:rv_bound}
\end{align}
for $f(e_{\rm c},\omega_{\rm c}, i_{\rm c}) \propto \sin^2 i_{\rm c}$
a geometric prefactor that depends on the argument of periastron
$\omega_{\rm c}$ and inclination $i_{\rm c}$ of the exterior companion
(\citealt{bailey_understanding_2019} Equation 23).  Since WASP-4b is
transiting, $f$ can be arbitrarily small, and both of the preceding
limits can be satisfied.

@@

For each planet, we asked: what fraction of the posterior for $t_{\rm
offset}$ is consistent with an offset worse than $81.6$ seconds?  For
WASP-4b, the answer is unsurprisingly 50\%.  For WASP-6b, the most
constraining object, about 1 sample in 2 million is consistent with
such a timing offset ($4.9\sigma$).  For WASP-18b, 1 in 103 samples
would be consistent with this timing offset ($2.3\sigma$), and in
WASP-46b, the limit is 1 in 49 samples ($2.0\sigma$).  For WASP-5b,
the predicted time is too imprecise to rule out timing offsets at the
necessary amplitude.  Multiplying the three independent probabilities
for WASP-5b, 6b, and 18b, we can rule out $t_{\rm offset} < -81.6\
{\rm seconds}$ at $6.4\sigma$, or about about 1 part in 11 billion.
